Using synchronized brain rhythms to bias memory-guided decisions

Version 1

Abstract

Abstract

Functional interactions between the prefrontal cortex and hippocampus, as revealed by strong oscillatory synchronization in the theta (6-11 Hz) frequency range, correlate with memory-guided decision-making. However, the degree to which this form of long-range synchronization influences memory-guided choice remains unknown. Here, we developed a brain machine interface that initiated trials on T-maze tasks based on the magnitude of prefrontal-hippocampal theta synchronization, then measured choice outcomes. Strong prefrontal-hippocampal theta synchrony was found to bias correct choices on both working memory dependent and cued-response tasks. Prefrontal-thalamic neural interactions increased with prefrontal-hippocampal synchrony and optogenetic activation of the ventral midline thalamus entrained prefrontal neural activity and prefrontal-hippocampal theta synchrony. Together, our results show that prefrontal-hippocampal theta synchronization leads to correct choices and strengthens prefrontal-thalamic dialogue. Our findings reveal new insights into the neural circuit dynamics underlying memory-guided choices and highlight a promising technique to potentiate cognitive processes or behavior via brain machine interfacing.

Introduction

Working memory, the ability to temporarily maintain and mentally manipulate information, is fundamental to cognition (Baddeley, 1986). This ability is known to require communication and neuronal processing across distributed brain regions and is conserved over mammalia (Goldman-Rakic, 1991; Sarnthein et al., 1998; Lee and Kesner, 2003; Winter and Stich, 2005; Wang and Cai, 2006; Eichenbaum, 2008; Fell and Axmacher, 2011; Christophel et al., 2017; Eichenbaum, 2017; Churchwell and Kesner, 2011; Spellman et al., 2015; Hallock et al., 2016; Ito et al., 2015; Bolkan et al., 2017; Ito et al., 2018; Maisson et al., 2018; Lugtmeijer et al., 2021). Long-range interactions are thought to be supported by the proper timing of action potentials (spikes), and brain rhythms are thought to act as a clocking mechanism to synchronize the timing of spike discharges (Fries, 2005; Buzsaki, 2006; Fell and Axmacher, 2011; Colgin, 2011; Fries, 2015). For example, hippocampal rhythms are coupled to the organization of hippocampal spiking activity in rats (O’Keefe and Recce, 1993), bats (Eliav et al., 2018), primates (Jutras et al., 2009), and humans (Qasim et al., 2021), although the exact frequency can vary over mammalia. The hypothesis that brain rhythms coordinate brain communication by synchronizing neuronal activity, known as “communication through coherence”, is just beginning to be experimentally tested (Fries, 2005; Buzsaki, 2006; Fell and Axmacher, 2011; Fries, 2015; Reinhart and Nguyen, 2019).

In rats, decades of research have shown that computations within, and communication between, the medial prefrontal cortex (mPFC) and hippocampus are required for spatial working memory (Dudchenko, 2001; Lee and Kesner, 2003; Wang and Cai, 2006; Horst and Laubach, 2009; Churchwell and Kesner, 2011; Hallock et al., 2013a). Recording studies specifically implicate theta synchrony within the mPFC-hippocampal network as a mechanism for mPFC-hippocampal communication. One metric of oscillatory synchrony, coherence, has been repeatedly correlated with memory-guided choices (Jones and Wilson, 2005; Benchenane et al., 2010; Sigurdsson et al., 2010; O’Neill et al., 2013; Hallock et al., 2016), but also with attention and task engagement (Guise and Shapiro, 2017; Bygrave et al., 2019). In a cornerstone experiment, Jones and Wilson (2005) showed that 4-12Hz mPFC-hippocampal coherence was stronger before rats made a correct choice when compared to a choice error or forced navigation on a spatial memory task. While foundational, this study treated the dependent variable (choice accuracy) as independent to test the effect of choice outcome on task performance. The same is true of other studies, where theta coherence (dependent variable) was shown to be positively correlated with choice accuracy (dependent variable; Benchenane et al., 2010; Hallock et al., 2016). Thus, did mPFC-hippocampal synchrony lead to correct choice outcomes, or co-occur with correct choices?

To understand the effect of mPFC-hippocampal theta coherence on choice, we developed a brain machine interface that tied theta (6-11Hz) coherence magnitude with task manipulation and measured choice outcomes on two different tasks. We are the first to show that strong mPFC-hippocampal theta coherence probabilistically led to more correct choices and that this result was true for a spatial working memory dependent task and a cued-response task. In follow-up experiments, we found that mPFC theta rhythms and mPFC-thalamic interactions increased with mPFC-hippocampal theta synchrony. Consistent with these results, optogenetic activation of the ventral midline thalamus, a structure known to coordinate mPFC-hippocampal interactions (Vertes, 2002; McKenna and Vertes, 2004; Gabbot et al., 2005; Vertes et al., 2006; Hoover and Vertes, 2007; Hoover and Vertes, 2012; Hallock et al., 2016; Dolleman-van der Weel et al., 2019; Griffin, 2021), reproduced findings seen during times of strong endogenous mPFC-hippocampal theta coherence.

Results

Development of a closed-loop brain machine interface for coherence-dependent task manipulation

Rats were implanted with stainless steel wires targeting the prelimbic and infralimbic subregions of the medial prefrontal cortex (mPFC) and CA1 of dorsal hippocampus (Figs 1A and 2A) to record local field potentials (LFPs). To support brain machine interfacing, we designed two independent loops, one processing the neural data in real time and the other controlling the automatic maze (Extended Fig. 1; Fig. 1A).

A brain machine interface that harnesses endogenous mPFC-hippocampal theta coherence on a working memory task.

A) Schematic of brain machine interfacing as rats performed a delayed alternation task on an automated T-maze. The delayed alternation task requires rats to alternate between left and right reward zones. Blue arrows and stars denote correct (rewarded) trajectories while red arrows and stars represent incorrect (unrewarded) trajectories. The rat was confined to the delay zone with three barriers. On a subset of delays, we computed mPFC-hippocampal theta coherence in real-time and trials were initiated contingent upon theta coherence magnitude. B) Frequency by coherence distribution calculated on data collected in real-time (N = 8 rats; 4 female, 4 male). For brain machine interfacing experiments, “theta coherence” was defined as the averaged coherence values between 6-11Hz. Data are represented as the mean +/− s.e.m. C) Thresholds for high and low magnitude coherence were estimated based on distributions of theta coherence values that were unique to individual rats (see Extended Fig. 3J and methods). N = 8 rats (4 female, 4 male).

High mPFC-hippocampal theta coherence can be used to enhance performance of a working memory dependent task

A) Left panel: Histology from a representative rat showing electrode tracks in the dorsal hippocampus (top) and mPFC (bottom). Right panel: Distribution of trial-types within a session. Within 10-trial blocks, 20% of trials were initiated based on high or low mPFC-hippocampal theta coherence, 20% of trials were yoked to the high/low coherence trials, and 60% were triggered following a random delay (5-30s). Yoked trials were identical in delay duration as high/low coherence trials, but triggered independent of coherence magnitude to control for the negative correlation between delay length and task performance (Extended Fig. 4A). B) Example LFP traces recorded during high and low coherence trials from three representative rats. The mPFC and hippocampal signals were used to compute theta coherence in real-time. C) Coherograms representing time around trial initiation (x-axis), coherence frequency (y-axis) and coherence magnitude, with warmer colors indicating higher coherence values. White arrows denote strong (top panel) and weak (bottom panel) theta coherence, as expected on trials triggered during high and low coherence states. Notice that on yoked trials, coherence was rather consistent before and after trial initiation, as expected for trials triggered independent of coherence magnitude. D) Relative to yoked trials, presenting choices to rats when mPFC-hippocampal theta coherence was high led to improved task performance (t(7) = 2.85, pp.c. = 0.0248). Trials contingent upon low magnitude theta coherence did not impact task performance compared to delay matched controls (t(7) = −0.26, pp.c. = 0.80; paired t-test). Follow-up statistical testing revealed that choice accuracy on high coherence trials was significantly greater than choice accuracy on random delays, consistent with our planned comparisons between high and yoked trials (t(7) = 6.12; p(x4) = 0.002). See Extended Table 1 for statistics. * p<0.05, **p<0.01. Stars (**) above bar graph denotes significance as measured from comparisons relative to random delay choice outcomes (black) and relative to 70% criterion (gray). Subscript “P.C.” indicates planned comparisons. Subscript “(x4)” indicates unplanned comparisons with Bonferroni corrected p-values for the number of unplanned tests performed.

By iteratively extracting signals into MATLAB from our acquisition system at systematically-increasing lags (25ms-300ms), we found that waiting 250ms before extracting new signals provided reliable streaming between the brain and MATLAB (Extended Fig. 2A-C). We then tested the impact of dataset sizes on the strength and the shape of the coherence distribution within the 4-12Hz range, in real time. By linearly increasing the amount of data being analyzed, and calculating coherence over 50 separate intervals from an example rat in real-time, we noticed that the dataset sizes strongly impacted the shape of the coherence distribution (Extended Fig. 2E-F), although the effect on coherence magnitude was less robust (Extended Fig. 2D). Since the strongest frequency (4-12Hz) plateaued at ∼8Hz when analyzing dataset sizes of 1.25s (Extended Fig. 2E), we chose to use 1.25s dataset sizes with 250ms steps forward in time (Extended Fig. 2G). This approach led to clear transitions between high and low magnitude theta coherence (Extended Fig. 2H) indicating that we were accurately tracking coherence in real-time.

Since brain machine interfacing handles data acquired in real time, we z-score transformed real-time detrended signals against a mean and standard deviation acquired during a 10 minute recording session that occurred on a previous day. We excluded epochs if >1% of the LFP signals were saturated with voltages exceeding 4 standard deviations. Since movement related artifacts often coincided with strong delta (1-4Hz) power (Extended Fig. 2I), we also excluded epochs if delta coherence was stronger than theta coherence. When combined, these approaches isolated coherence distributions with clear theta synchrony (6-11Hz; Fig. 1B) and high consistency across rats (Extended Fig. 2I).

Once these algorithms were defined (Extended Figs 1-3), we then trained rats to perform a delayed alternation task in a T-maze until they performed at 70% choice accuracy for two consecutive days. On this spatial working memory task, rats are rewarded for alternating between left and right reward zones, and sequestered at the base of the maze before each choice (Fig. 1A). The ability of this task to tax working memory was validated by measuring the impact of delay duration on choice outcome. Consistent with the use of delayed-response tasks across species (Dudchenko, 2001; Goldman-Rakic, 1991; Eichenbaum, 2008), longer delay durations led to lower choice accuracy (Extended Fig. 4A)

High mPFC-hippocampal theta coherence trials are gated by prefrontal theta rhythms and lead to heightened pre-choice synchrony

A) Prefrontal and hippocampal power spectra during the high and low coherence epochs used for brain machine interfacing (Fig. 1 and 2). B) Prefrontal theta power (6-9Hz) was significantly greater during high coherence epochs relative to low coherence epochs (t(7) = 3.66, ci = 0.067 to 0.312, padj(x2) = 0.016). Hippocampal theta power was stronger on high coherence compared to low coherence trials (t(7) = 2.36, ci = −0.0003 to 0.254, p = 0.05). C) The frequency of prefrontal and hippocampal theta oscillations was significantly higher during high coherence states relative to low coherence states (PFC: t(7) = 5.35, padj(x2) = 0.002, ci = 0.55 to 1.42; hippocampus: t(7) = 3.34, padj(x2) = 0.025, ci = 0.17 to 0.98). Theta frequency was measured by identifying the frequency corresponding to maximum theta power. D) Hippocampal-to-prefrontal theta directionality was significantly stronger during high theta coherence states relative to low theta coherence states (t(7) = 3.9, ci = [0.11 to 0.46], padj(x3) = 0.018). No significant effect was observed in the prefrontal-hippocampal direction (t(7) = 1.8, p = 0.11), nor when we compared HPC->PFC vs PFC->HPC (t(7) = 2.6, padj(x3) = 0.1). For panels B-D, data are represented as the mean +/− s.e.m. across 8 rats. *p<0.05, **p<0.01, paired t-tests with Bonferroni p-value corrections when p<0.05. Difference scores were tested against a null of 0. E) LFP signals (jittered for visualization) were extracted from 2s before choice point entry (as defined by infrared beam breaks) and 0.5s afterwards. Bar graphs show that the average time to choice-entry for high coherence and low coherence trials was between 1.6-2.1s and did not differ between trial-types (t(7) = 2.0, p 0.08). F) Averaged coherograms (N = 8 rats) showing coherence as a function of frequency and time surrounding choice point entry. G) Difference of the coherograms shown in F. White arrows point to initial 6-9Hz synchronization at −2s which approximates trial onset (see bar graph in E), and a second time point of heightened theta synchrony before choice entry. H) Normalized difference scores representing theta coherence as a function of time. Theta coherence at choice-entry was significantly stronger on trials triggered by high coherence relative to trials triggered during low coherence (see Extended Table 2). Magenta lines denote p<0.05 after Benjaminin Hochberg corrections.

mPFC-hippocampal theta coherence was harnessed to enhance the performance of a cue-guided decision making task

A) Schematic of the conditional discrimination task. Wooden or mesh floor inserts were used to guide choice behavior. Rats were randomly assigned to insert-reward contingencies. Like the brain machine interfacing experiment on the delayed alternation task, trials were initiated when rats were sequestered in the delay zone. B) Example histology from a representative rat showing electrode placements in the hippocampus and mPFC. C) Trials initiated during high mPFC-hippocampal theta coherence states led to better task performance when compared to yoked control trials (t(15) = 2.23, ci = 0.29 to 12.87, p(p.c.) = 0.04) or when compared to trials triggered following a random delay (t(15) = 3.8, ci = 4.7 to 16.6, p(x2) = 0.003). There was no difference in choice outcome following yoked and random delay trials (t(15) = 1.0, ci = −4.5 to 12.7, p(x2) = 0.33). *p<0.05. **p<0.01. Subscript on p-values show if comparisons were planned (‘p.c.’) or corrected for multiple comparisons (‘x2’). Data are represented as the mean ± s.e.m.

During training sessions, thousands of theta coherence values were calculated during the delay phases, and distributions of mean theta coherence estimates were created (Extended Fig. 2J). Using these distributions, we defined weak theta coherence as 1std below the mean, and strong theta coherence as 1std above the mean of all theta coherence values. Therefore, each rat had a unique numerical value defining states of strong and weak theta coherence, which we could then use as thresholds to initiate trials on the automated maze (Fig. 1A; Extended Fig. 3).

Strong prefrontal-hippocampal theta coherence leads to correct choices on a spatial working memory task

Based on multiple reports, mPFC-hippocampal theta coherence is positively correlated with memory-guided decision making (Jones and Wilson, 2005; Benchenane et al., 2010; Hallock et al., 2016). The results from such experiments depend on the comparison of two dependent measurements (choice accuracy and theta coherence), but whether theta coherence can be used with task manipulation (independent variable) to bias choice accuracy (dependent variable) remains unexplored. To test this idea, we implemented the algorithms described above with an automatic maze to control trial initiations. During experimentation, our brain machine interface was activated as rats occupied the delay zone and rats were presented with various trial types within a given session as follows. A small proportion of trials were initiated when mPFC-hippocampal theta coherence was above the strong theta coherence threshold (∼10% of trials) or below the weak theta coherence threshold (∼10% of trials) (Fig. 2A and 2B). Since increasing delay durations lead to worse task performance (Extended Fig. 4A), rats also experienced trials that were yoked to high and low coherence trials via identical delay durations. For example, if trial N was a high coherence trial, our algorithm logged the duration spent in the delay zone to be presented back to the rat within a 10-trial block. Thus, initiation of yoked trials was independent of the strength of theta coherence (Fig. 2C) and by comparing choice accuracy on strong/weak coherence trials to that on yoked trials, we were able to rule out the possible confounding variable of working memory load on choice accuracy.

We predicted that, relative to yoked control trials, trials presented during states of strong mPFC-hippocampal theta coherence would be more likely to be correct and trials presented during states of weak mPFC-hippocampal theta coherence would be more likely to be incorrect. Consistent with our first prediction, presenting trials during elevated states of mPFC-hippocampal theta coherence improved choice accuracy (Fig. 2D). However, choice accuracy on trials presented during states of low mPFC-hippocampal theta coherence did not differ from choice accuracy on yoked control trials, indicating that naturally occurring weak theta synchronization does not impair choice outcomes. This latter result may be explained by a default mode, such as parallel processing (Wang and Cai, 2006; Churchwell and Kesner, 2011)

We then examined various measurements of overt behavior to test if behaviors differed between coherence-triggered trials and yoked trials. First, we examined the amount of time spent until rats made a choice, defined as the amount of time from the point at which a trial is initiated until rats passed the infrared beam that triggers the reward dispenser (Extended Fig. 1). While we found no difference in time-to-choice between high coherence trials and yoked trials, there was a trending difference between low and yoked trials (Extended Fig. 4B). Using an analysis to test head-movement complexity (IdPhi; Papale et al., 2012; Redish, 2016), we found no differences between high coherence trials and yoked trials, but did observe less head-movement complexity on low coherence trials relative to yoked trials (Extended Fig. 4C). Next, we analyzed total distance traveled in the epoch used to trigger trials during high and low coherence states (last 1.25s before trial initiation). Since the amount of time was always consistent (1.25s), this approach is a proxy for speed, an indirect correlate of theta frequency (Kropff et al., 2021). We found no differences in movement behavior between coherence trials and yoked trials (Extended Fig. 4D). Finally, we found that rats spent similar amounts of time in the delay zone during high and low coherence trials (Extended Fig. 4E). These analyses show that high coherence trials could be used to promote correct choices in the absence of overt differences in behavior between trial types, indicating that mPFC-hippocampal theta coherence may play a causal role in memory-guided choices.

Trials initiated by strong prefrontal-hippocampal theta coherence are characterized by prominent prefrontal theta rhythms and heightened pre-choice prefrontal-hippocampal synchrony

Next, we performed offline data analysis to understand the neural dynamics occurring during the high coherence states that improved spatial working memory task performance. First, we noticed that theta rhythms were better characterized by changes within the 6-9Hz range (Fig. 3A) and as such, offline analyses focused on this narrow band. Relative to low coherence states, mPFC theta rhythms were stronger during high coherence states (Fig. 3A-3B; see Fig. 2B for example LFP traces). Hippocampal theta rhythms only exhibited a modest elevation in theta power relative to low coherence states. With respect to theta frequency, both mPFC and hippocampal theta rhythms were shifted toward higher frequencies during high coherence states relative to low coherence states, with mPFC theta rhythms approaching 7 Hz and hippocampal rhythms approaching 8 Hz (Fig. 3C). We then analyzed whether these signals exhibited evidence of directionality, the ability for one signal to predict another signal as measured by Granger causality analysis (Cohen, 2014). Relative to low coherence states, high coherence states were characterized by stronger hippocampal-to-mPFC theta directionality (Fig. 3D). Thus, the high theta coherence states used to trigger spatial working memory trials were characterized by strong mPFC theta rhythms and hippocampal-to-mPFC theta directionality.

Even though the delay zone was physically close to the choice point (∼30cm), we wondered whether strong mPFC-hippocampal theta coherence trials impacted synchronization during the goal-arm choice. Therefore, we defined choice-point entry as the infrared beam break immediately preceding the choice (Extended Fig. 1). On average, rats took 1.6s and 2.1s to reach this infrared beam from trial initiation on low and high coherence trials, respectively. No significant difference in time-to-choice was observed between high and low coherence trials (Fig. 3E). Thus, we extracted LFPs from −2s to +0.5s surrounding choice-entry (Fig. 3E), and calculated coherence over time and frequency (Fig. 3F). A normalized difference score was calculated from the resultant coherograms (high-low/high+low), revealing a clear difference in theta coherence magnitude between high and low coherence trials as rats approached the choice zone (Fig. 3G). As expected, high coherence trials showed significantly stronger synchronization at −2s, an approximate for trial initiation (Fig. 3H). Interestingly, after the 2s time-point, theta coherence between high and low coherence trials became more similar, but once again differed at ∼0.5s pre-choice (Fig. 3H). This latter result shows that strong mPFC-hippocampal theta coherence during the delay was maintained up until the time of goal arm choice.

Given that mPFC-hippocampal theta coherence fluctuated in a periodical manner (Extended Fig. 5B), we wondered how predictive past values of mPFC-hippocampal theta coherence were of future values. Using previously collected data (Hallock et al., 2016), we extracted mPFC-hippocampal theta coherence epochs across the duration of a 30s delay on the delayed alternation task from 3 rats (N = 22 sessions; Extended Figs 5A and 6A). We performed an autocorrelation analysis on theta coherence values on a trial-by-trial basis, then compared the results to a temporally shuffled theta coherence distribution. Since we performed a moving window approach (1.25s in 250ms steps), comparisons between real and temporally shuffled coherence estimates were only included after 5 lags (lag #4 relative to 0; Extended Fig. 5B). Consistent with the periodical nature of mPFC-hippocampal theta coherence epochs, there was a significant correlation between past and future coherence estimates, even when epochs shared no overlap in data (Extended Fig. 5C). Thus, mPFC-hippocampal theta coherence is dynamic and periodical.

Prefrontal-hippocampal theta synchronization modulates prefrontal-thalamic interactions

A) LFPs were recorded from the mPFC, VMT and hippocampal of 3 rats (N = 22 sessions). Right panel shows triple site recordings taken from a representative rat. Green box shows example tetrode tracks from the mPFC. B) High and low mPFC-hippocampal theta coherence epochs were identified, and LFP from the VMT was extracted. The data shown are collapsed across high or low coherence epochs. C) Frequency by coherence plots from the mPFC (top panel), VMT (middle panel), and hippocampus (bottom panel). Compare these data to Fig. 3. D) Normalized difference scores comparing theta (6-9Hz) power between high and low coherence epochs. There was a main effect of brain region on the coherence difference score (F(2,65) = 20.8; p < 0.001; one-way ANOVA) with each brain area showing higher theta power during high coherence states relative to low coherence states (PFC: p < 0.001; VMT; p < 0.001; HPC: p < 0.001; see Extended Table 3). E) Theta coherence for mPFC-VMT and VMT-HPC was estimated during high and low mPFC-hippocampal theta coherence states. F) mPFC-VMT and VMT-HPC theta coherence was stronger during high when compared to low mPFC-hippocampal theta coherence states. mPFC-VMT theta coherence changed more drastically with mPFC-hippocampal theta coherence magnitude (mPFC-VMT: p < 0.001; VMT-HPC: p < 0.001; mPFC-VMT vs VMT-HPC: p < 0.001; see Extended Table 4). G) Multivariate granger prediction analysis. Left panel shows VMT-HPC theta directionality. Middle panel shows mPFC-VMT theta directionality. Right panel shows mPFC-hippocampal theta directionality. Granger prediction in the mPFC- to-VMT direction was more sensitive to mPFC-hippocampal theta coherence magnitude when compared to granger prediction in the VMT-to-mPFC direction (statistics in Extended Table 5). H) Top panel shows hippocampal LFP (1-sec) and example spikes from an mPFC neuron with significant spike-theta entrainment. Middle panel shows polar plots of the unit in the top panel. Histogram represents the distribution of spike-phase values with the mean result length vector shown as a white bar in the center. Bottom panel shows spike-field coherence for the same neuron. I) Difference score (high-low/high+low) of bootstrapped MRL and Rayleighs Z-statistic for each neuron as a function of hippocampal or VMT theta. No significant differences were found between high and low mPFC-hippocampal theta coherence states. J) Spike-field coherence, represented as a difference score. No effects survived p-value correction. Arrow points to a numerical increase to spike-field coherence at hippocampal 4-6Hz. K) Percentage of significantly modulated mPFC units to VMT theta and hippocampal theta. *p<0.05. **p<0.01. Data are represented as the mean ± s.e.m.

Optogenetic activation of the ventral midline thalamus produces prefrontal theta and dynamically modulates prefrontal-hippocampal theta coherence

A) Top panel, Schematic demonstrating recordings from the mPFC and hippocampus with optogenetic activation of the VMT. Middle panel, example histological confirmation of fiber implant and viral expression targeting the VMT. Bottom panel. Viral expression at similar viral injection coordinates. Notice that all rats showed overlap in viral expression in the nucleus reuniens (brain section overlay from Paxinos and Watson, 2006). B) Optogenetic activation of the VMT at 7Hz produced prefrontal theta rhythms (N = 83 blue, 88 red laser events). C) Power spectrum from blue and red laser stimulation events averaged over recording channels. Columns represent recording channels per shank, while rows represent shank number and the corresponding medial-lateral placement in the mPFC (B). D and E) Data from rat 21-42 (N = 108 blue, 104 red laser events). D) Top panel shows raw LFP traces, middle panel shows theta filtered traces (6-9Hz), while the bottom panel shows theta coherence as a function of time. Yellow box shows stimulation the event. E) VMT activation increased mPFC and hippocampal theta, while reducing other frequencies of the mPFC and hippocampal LFP. VMT activation decreased mPFC-hippocampal theta coherence (bottom panel). F and G) Data from 21-43 (N = 113 blue, 101 red laser events), like those results from 21-42. Notice that VMT stimulation robustly increased mPFC theta power, while reducing mPFC-hippocampal theta coherence. H-K) Data from 21-42 (N = 63 blue, 104 red laser events) and 21-43 (N = 113 blue, 101 red laser events) after excluding the first 0.5s, where visual observations revealed a brief decoupling of mPFC and hippocampal theta oscillations. Both rats showed increased mPFC-hippocampal theta coherence. Lines denote p<0.05 following Benjamini-Hochberg p-value corrections for two-sample t-tests. Data are represented as the mean ± s.e.m.

In our brain machine interfacing experiments, trials were initiated when mPFC-hippocampal theta coherence was strong or weak. States of strong mPFC-hippocampal theta coherence increased the probability of a correct choice, while increasing synchronization during task performance. However, when we examined the frequency of strong mPFC-hippocampal theta coherence events when the delay phase was fixed and predictable, strong mPFC-hippocampal theta coherence events did not predict trial initiation (Extended Fig. 5D). Therefore, strong mPFC-hippocampal theta coherence led to improved task performance through mechanisms specific to choice.

Prefrontal-hippocampal theta coherence states lead to correct choices on a cue-guided task

Our findings from Fig. 2 show that mPFC-hippocampal theta coherence leads to correct spatial working memory-guided choices. We next wondered if this effect was specific to spatial working memory and specifically tested whether strong mPFC-hippocampal theta coherence events were optimal for choices on a task where rats must attend to external stimuli to guide decision making. Rats (N = 3; 1 male, 2 female) were implanted with wires targeting the mPFC and hippocampus (Fig. 4B) and were trained to perform a conditional discrimination task where a floor insert dictated choice outcome (e.g. a wooden floor insert might dictate a left choice, while a mesh insert might dictate a right choice; Fig. 4A). This task is similar in difficulty to the delayed alternation task, but requires the striatum, rather than the hippocampus to perform (Griffin, 2012; Hallock et al., 2013a). Likewise, past research showed that inactivation of the mPFC or the ventral midline thalamus did not disrupt conditional discrimination task performance in well-trained rats (Hallock et al., 2013b, Shaw et al., 2013), indicating that the mPFC-hippocampal network is not required for conditional discrimination task performance. Therefore, we predicted that strong mPFC-hippocampal theta coherence would not improve choice outcomes on this cue-guided task.

We collected 35 sessions, of which 16 sessions (7 sessions from 21-48 [male]; 4 sessions from 21-49 [female]; and 5 sessions from 21-55 [female]) met criterion for performance of >70%, alternation of <70%, and a contribution of at least 4 trials. Unexpectedly, we found that initiation of trials during strong mPFC-hippocampal theta coherence enhanced choice accuracy on the conditional discrimination task (Fig. 4C). This finding was surprising given that mPFC-hippocampal theta coherence did not previously correlate with choice outcomes on the conditional discrimination task (Hallock et al., 2016), but consistent with increased mPFC-hippocampal theta coherence on a different cue-guided paradigm (Benchenane et al., 2010). Most importantly, these results show that strong mPFC-hippocampal theta coherence is optimal for decision making behavior regardless of whether working memory and mPFC/hippocampal function is necessary to perform a task.

Prefrontal-thalamo-hippocampal network dynamics vary with prefrontal-hippocampal synchronization

So far, we have shown that initiating trials when mPFC-hippocampal theta synchrony is strong leads to correct memory-guided choices. What are the mechanisms supporting strong mPFC-hippocampal theta synchrony leading to improved choice accuracy? Past research showed that mPFC-hippocampal theta synchrony during choice was supported by the ventral midline thalamus (VMT; Hallock et al., 2016). The VMT is anatomically connected with the mPFC and hippocampus (Sesack et al., 1989; Vertes, 2002; McKenna and Vertes, 2004; Vertes, 2006; Hoover and Vertes, 2007; Hoover and Vertes, 2012), providing a source of glutamatergic excitation to both structures (Dolleman-van der Weel et al., 2019). Therefore, we wondered how mPFC-VMT and VMT-hippocampal interactions varied with mPFC-hippocampal theta synchronization.

To probe this question, we examined datasets with simultaneous mPFC, VMT, and dorsal hippocampus recordings from 3 rats performing a spatial working memory task (N = 22/28 sessions; Fig. 5A; Extended Fig. 6B; Stout and Griffin, 2020). We extracted neural data as rats occupied the delay zone, then defined and detected epochs of strong and weak mPFC-hippocampal theta coherence offline (Fig. 5B; Extended Fig. 6A-B). Corroborating the findings from our brain machine interfacing experiment (Figs 2 and 3), high theta coherence states were characterized by strong 6-9Hz theta rhythms in the mPFC (Figs 5C and 5D). Intriguingly, the magnitude change of theta power between high and low coherence states was strongest in the mPFC, followed by the VMT, then the hippocampus (Fig. 5D). Relative to low coherence epochs, the mPFC was differentially and simultaneously synchronized to the VMT and hippocampus during high coherence states (Fig. 5E). Moreover, high coherence states were characterized by a stronger change in neural synchronization between the mPFC and VMT, relative to the VMT and hippocampus (Fig. 5F). This latter result suggested that mPFC-VMT interactions may be particularly sensitive to mPFC-hippocampal synchronization. In support of this conclusion, multivariate granger prediction revealed that mPFC-VMT directionality was elevated during strong relative to weak mPFC-hippocampal theta coherence states (Fig. 5G; middle panel). mPFC-hippocampal directionality was also modulated by mPFC-hippocampal theta coherence magnitude. However, directionality between the VMT and hippocampus was minimally impacted by the magnitude of mPFC-hippocampal theta coherence (Fig. 5G).

Lastly, we examined whether mPFC spike-LFP synchrony was impacted by mPFC-hippocampal theta coherence. Spike-phase entrainment was used to quantify the non-uniformity of spike-phase distributions at theta to measure theta phase locking, and spike-field coherence was used to understand the correlation between spikes and LFP across frequencies (Fig. 5H). Out of 126 mPFC neurons, 46 neurons met criterion for inclusion (see methods). When comparing strong to weak mPFC-hippocampal theta coherence states, there were no significant differences to theta phase entrainment (Fig. 5I) nor to spike-field coherence (Fig. 5J) of mPFC spikes to VMT and hippocampal theta.

We then wondered if strong mPFC-hippocampal theta coherence states modulated the spike timing of a select group of mPFC neurons. During strong mPFC-hippocampal theta coherence states, 8.9% and 7% of mPFC neurons were modulated by hippocampal theta and VMT theta, respectively. This contrasted with weak mPFC-hippocampal theta coherence states, where 4.4% and 2.3% of mPFC neurons were significantly modulated by hippocampal and VMT theta, respectively (Fig. 5K). These findings indicate that the magnitude of mPFC-hippocampal theta synchronization, which was used to enhance memory guided choice (Figs 2 and 3) and known to modulate mPFC neuronal discharge (Benchenane et al., 2010; Fig. 5K), is also characterized by an increased proportion of mPFC neurons phase locking to VMT and hippocampal theta.

Optogenetic activation of the VMT dynamically regulates prefrontal-hippocampal theta rhythms

Next, we examined whether artificial theta frequency stimulation of the VMT was sufficient to produce synchronized theta rhythms between the mPFC and hippocampus. To investigate this question, we injected the VMT with AAV5-hSyn-ChR2-eYPF to create and embed channelrhodopsin2 at the membrane of VMT neurons, a proton pump that promotes excitation of neurons with blue light stimulation (450nm). After 4-6 weeks of recovery to allow for viral expression, we pulsed a blue laser at 7-8Hz (2s on 2-8s off) targeting the VMT while recording from the mPFC (N = 3 rats) and the hippocampus (N = 2 rats; Fig. 6A). As a within-subject control, we also stimulated the VMT with a red laser (638nm). Stimulation with red and blue lasers were randomly interleaved within a recording session.

Optogenetic stimulation of the VMT between 7-8Hz produced a large negative deflection in the mPFC voltage and enhanced mPFC theta rhythms across all rats (Fig. 6B, D, F, H, and J). Using a 64ch silicon probe targeting mPFC lamina, optogenetic activation of the VMT enhanced the power of mPFC theta oscillations across all recording channels (Fig. 6B-C). This result demonstrated the clear impact of VMT activation on all lamina of the mPFC.

The effect of VMT activation on mPFC-hippocampal theta coherence was initially disruptive (Fig. 6D-G), which led us to visually inspect the raw and theta filtered signals. This approach revealed that upon VMT stimulation, some trials showed a desynchronization of mPFC-hippocampal theta rhythms, followed by synchrony (Fig. 6H and J). We therefore excluded the first 0.5s of stimulation, focusing on a 1s window of VMT stimulation. In rat 21-42, we noticed instances where VMT stimulation had a rather small or no clear effect on mPFC theta, which was unexpected based on the data from 21-45 and 21-43. Given that viral injections were focused on one rostro-caudal location, rather than being distributed across the entirety of the ventral midline thalamus (see Dolleman Van-der Weel et al., 2019 for a review), we performed K-means clustering (N = 2 clusters) to identifying times when VMT stimulation increased mPFC theta power. When combined with the data from 21-43, we found that VMT activation increased mPFC-hippocampal theta coherence (Fig. 6I and 6K). Taken together, VMT stimulation dynamically impacted mPFC-hippocampal theta coherence, altered hippocampal oscillations, but drastically influenced mPFC theta rhythms.

Discussion

Past research has found that mPFC-hippocampal theta coherence was stronger when memory was used to guide choices (Jones and Wilson, 2005; Benchenane et al., 2010; Sigurdsson et al., 2010; O’Neill et al., 2013; Hallock et al., 2016). However, these results required the comparison of two dependent measurements that have since been shown to be inter-related (Benchenane et al., 2010; Hallock et al., 2016). Thus, did mPFC-hippocampal theta coherence lead to, or coincide with, correct choice outcomes?

To answer this question, we implemented brain machine interfacing techniques to incorporate mPFC-hippocampal theta coherence as an independent variable (i.e. task control), and demonstrated for the first time that strong mPFC-hippocampal theta coherence enhances memory-guided choice. While we expected this form of long-range theta synchronization to be particularly useful when spatial working memory was used to guide decision-making, we also observed that mPFC-hippocampal theta coherence enhanced the performance of a task that does not require the mPFC, VMT, nor hippocampus for successful performance (Hallock et al., 2013a; Hallock et al., 2013b, Shaw et al., 2013). Surprisingly, this latter result indicates that structures deemed unnecessary for a cognitive process can still contribute rather significantly. Based on the communication through coherence hypothesis, we suspect that strong mPFC-hippocampal theta synchronization improved memory-guided choice via increasing the probability of neuronal coordination.

To further test the communication through coherence hypothesis, we extracted strong and weak mPFC-hippocampal theta synchronization events, then compared mPFC-thalamo and hippocampal-thalamo interactions. For these analyses, we focused on the ventral midline thalamus (VMT), a structure that is bidirectionally connected with the mPFC and hippocampus and supports mPFC-hippocampal neuronal interactions (Vertes, 2002; McKenna and Vertes, 2004; Vertes et al., 2006; Hoover and Vertes, 2007; Hoover and Vertes, 2012; Ito et al., 2015; Hallock et al., 2016; Stout et al., 2022). mPFC spike times and mPFC theta oscillations synchronized more strongly with VMT theta rhythms when mPFC and hippocampal theta were strongly coherent. Likewise, theta rhythms between the VMT and hippocampus were coupled more strongly when mPFC and hippocampal theta rhythms were synchronized. Given that strong mPFC-hippocampal theta coherence events were previously shown to organize cell assemblies within the mPFC (Benchenane et al., 2010), our results indicate that these events enhanced memory-guided choice by structuring neuronal interactions. Given that mPFC-VMT neural interactions varied with mPFC-hippocampal theta synchronization, and that optogenetic theta stimulation of the VMT modulated mPFC theta and mPFC-hippocampal theta synchrony, our results point towards cortico-thalamic dialogue as a central component of mPFC-hippocampal theta synchronization. Importantly, this latter assertion is supported by anatomy, as the mPFC receives no direct projections from the dorsal hippocampus (Jay and Witter, 1991; Hoover and Vertes, 2007), but influences hippocampal neuronal activity via the thalamus (Ito et al., 2015). We suspect that the VMT may coordinate mPFC-hippocampal neural interactions through cortico-thalamo-cortical looping mechanisms, as the VMT projects directly to entorhinal cortex neurons that target the CA1 (Wouterlood, 1991) and modulates CA1 neurons with concurrent cortical activation (Dolleman-Van der Weel, 2017). Consistent with this hypothesis, mediodorsal thalamus is known to sustain mPFC neuronal activity (Bolkan et al., 2017; Schmitt et al., 2017), and the VMT supports mPFC firing and mPFC-hippocampal synchronization (Hallock et al., 2016; Jayachandran, Viena et al., 2023). We therefore hypothesize that states of strong mPFC-hippocampal theta synchronization probabilistically increased correct decisions through cortico-thalamic communication.

Importantly, our findings provide a promising avenue to rescue cognitive deficits in clinical populations. Consistent with our work, a recent study found that inducing states of theta synchrony between frontal and temporal regions via transcranial alternating-current stimulation, rescued age-related memory impairments in human participants (Reinhart and Nguyen, 2019). Our findings suggest that tapping into pre-existing neural dynamics holds significant promise for improving memory. We hypothesize that non-invasive stimulation techniques prior to therapy, paired with synchrony-dependent attention or working memory practice via brain machine interfacing, could pose a viable intervention to improve cognitive deficits. In closing, the use of brain machine interfacing holds significant promise for clinical and neuroscientific advance.

eLife Assessment

This study enhances our understanding of the relationship between cortico-hippocampal interactions and behavioral performance. Using an inter-areal coherence metric to gate trial initiation in real time, the authors provide solid evidence that links high hippocampal-prefrontal theta coherence to correct performance on spatial working memory and cue-guided decision-making tasks. Although reviewers agreed that the results do not demonstrate causality between hippocampal-prefrontal synchrony and behavioral performance, the findings are viewed as important given their potential implications for brain-machine interface applications in humans.

Reviewer #1 (Public Review):

Summary:

Information transfer between the hippocampus and prefrontal cortex is thought to be critical for spatial working memory, but most of the prior evidence for this hypothesis is correlational. This study attempts to test this causally by linking trial start times to theta-band coherence between these two structures. The authors find that trials initiated during periods of high coherence led to a dramatic improvement in performance. This applied not only to a spatial working memory task, but also to a cue-guided navigation task, suggesting that coherence in these regions may be a signature of a heightened attentional or preparatory state. The authors supplement this behavioral result with electrophysiological recordings to test whether the ventral midline thalamus is likely to mediate hippocampal-prefrontal coherence.

Strengths:

This study demonstrates a striking behavioral effect; by changing the moment at which a trial is initiated, performance on a spatial working memory task improves dramatically, from around 80% correct to over 90% correct. A smaller but nonetheless robust increase in accuracy was also seen in a texture discrimination task. Therefore, prefrontal-hippocampal synchronization in the theta band may not only be important for spatial navigation but may also be associated with improved performance in a range of tasks. If these results can be replicated using noninvasive EEG, it would open up a powerful avenue for modulating human behavior.

Weaknesses:

Ventral midline thalamic nuclei, such as reuniens, have reciprocal projections to both the prefrontal cortex and hippocampus and are therefore well-situated to mediate theta-band interactions between these structures. However, alternative mechanisms cannot be ruled out by the results of this study. For example, theta rhythms are globally coherent across the rodent hippocampus, and the ventral hippocampus projects directly to the prefrontal cortex. Theta propagation may depend on this pathway, and may only be passively inherited by VMT.

The optogenetic manipulations are intended to show that theta in VMT propagates to PFC and also affects HPC-PFC coherence. However, the "theta" induced by driving thalamic neurons at 7 Hz is extremely artificial. To demonstrate that VMT is causally involved in coordinating activity across HPC and PFC, it would have been better to optogenetically inhibit, rather than excite, these nuclei. If the authors were able to show that the natural occurrence of theta in PFC depends on activity in VMT, that would be a much more convincing test of their hypothesis.

Reviewer #2 (Public Review):

A number of previous reports have demonstrated that theta synchrony between the hippocampus (HPC) and prefrontal cortex (PFC) is associated with correct performance on spatial working memory tasks. The main goal of the current study is to examine this relationship by initiating trials either randomly (as is typically done) or during periods of high or low PFC-HPC coherence. To this end, they develop a 'brain-machine interface' (BMI) that provides real-time estimates of PFC-HPC theta coherence, which are then used to control trial onset using an automated figure-eight maze. Their main finding is that choice accuracy is significantly higher on trials initiated when theta coherence is high whereas performance on low coherence trials does not differ from randomly initiated control trials. They also observe a similar result using a non-working memory task in the same maze.

Overall the main experiments (Figures 1-4) are well designed and the BMI approach is convincingly validated. There are also appropriate controls and analyses to rule out behavioral confounds and the results clearly presented. The idea of triggering trial onset based on brain activity is an interesting idea and helps to examine how extremes in the distribution of brain states are associated with behavior, something that might be more difficult to examine if trials are initiated randomly. As such, the BMI is an interesting approach for studying the neuronal basis of behavior that could potentially be applied beyond the particular field of the study, something the authors could perhaps have elaborated on more.

That being said, although the authors have elegantly revealed an association between PFC-HPC theta synchrony and behavior using their BMI approach, it is not apparent whether these results add substantially to previous reports of similar associations, including from the author's own work. The authors sometimes seem to claim that they do; for example, in the discussion, after describing previous studies that reported an association between PFC-HPC theta synchrony and behavior, they raise the reasonable question "did mPFC-hippocampal theta coherence lead to, or coincide with, correct choice outcomes?" What they subsequently write gives the impression that their study has somehow addressed the question, whereas in fact their results still leave this question open. For example, it is entirely possible that during high-coherence trials an unobserved neural process is influencing both coherence and task performance. The authors could have made a more convincing case as to why their correlative results go beyond similar findings from previous studies, perhaps by including additional analysis to strengthen their case.

Sometimes, the authors also seem to suggest that their results establish a causal relationship between synchrony and behavior, for example when they say that they have "demonstrated for the first time that strong mPFC-hippocampal theta coherence ENHANCES memory-guided choice" (line 557, my emphasis). However, causal manipulations of PFC-HPC synchrony would be required to make such claims. I am not suggesting that the lack of such data is necessarily a weakness of the study, only that causal claims are not supported by the author's results.

In addition to the behavioral results described above, the authors also examine how HPC-PFC synchrony modulates synchrony with the ventromedial thalamus (VMT; Figure 5) and how optogenetic modulation of the VMT influences PFC-HPC synchrony (Figure 6). However, these results feel somewhat more preliminary and their relationship to the other findings in the manuscript is not always clear. For example, given that the authors demonstrate that "Prefrontal-hippocampal theta synchronization modulates prefrontal-thalamic interactions" (Figure 5) I would rather have expected the authors to manipulate HPC and/or PFC and see how this affects VMT in Figure 6. It is also difficult to draw strong conclusions about the effects of optogenetic VMT stimulation since the results presented by the authors come from only 2 rats (Figure 6D-K) and therefore feel somewhat anecdotal. I could also not find any statistical test supporting the increase in the proportion of phase-locked neurons during high theta states shown in Figure 5K.

Reviewer #3 (Public Review):

Stout et al investigate the link between prefrontal-hippocampal (PFC-HPC) theta-band coherence and accurate decision-making in spatial decision-making tasks. Previous studies show that PFC-HPC theta coherence positively correlates with learning of these tasks and correct decisions but the nature of this relation relies on correlations that cannot show whether coherence leads, coincides, or is a consequence of decision making. To investigate more precisely this link, the authors devise a novel paradigm. In this paradigm, the rat is blocked during a delay period preceding its choice and they control on a trial-by-trial basis the level of PFC-HPC theta coherence prior to the decision by allowing the rat to access the choice point only at a time when coherence reaches above or below a threshold. The design of the paradigm is very nicely controlled thanks in particular to the trial-by-trial matching of delay period durations which is crucial for the working memory task. Moreover, the behaviour of the animal is strongly similar during high and low coherence periods which bolsters the specificity of the author's interpretation. Thanks to this approach, the authors clearly demonstrate that high theta coherence prior to choice-making is strongly predictive of better decision-making both in working memory and a cue-guided version of the task.

This novel paradigm provides an improvement in the level of experimental control to analyse the coherence/choice link but the exact interpretation of the results it yields is not entirely clear in the current manuscript. Using the PFC-HPC theta coherence during the delay period to the gate when the rat accesses the choice zone clearly separates this coherence from the behavioural decision itself. This provides convincing support for the idea that PFC-HPC theta coherence prior to the behavioural decision is related to correct decision-making and is not simply temporally coincidental or a consequence of the decision output. It does not however substantially increase the weight of evidence in favour of a causal link between theta coherence and correct decisions as suggested in the abstract ("PFC-HPC theta synchronization leads to correct choices"). Indeed, the paradigm does not de-correlate PFC-HPC theta synchronization from other neurophysiological variables such as neuromodulation, arousal, synchrony with other areas, etc that could be playing the true causal role in modulating decision-making.

The question of the link between the manuscript's findings and causal involvement of PFC-HPC dialogue is interestingly highlighted by the author's unexpected result showing that their paradigm reveals a link between theta coherence even in a sensory-driven version of the task. As the authors point out, results based on muscimol inhibition have shown that neither PFC nor HPC, nor the ventral midline thalamus, that mediates communication between the two, are involved in this task. This raises the question of why coherence between two areas is predictive of choice accuracy when neither area appears to be causally involved. The manuscript does not discuss the possibility that these results could imply that theta coherence is not in fact a good causal indicator. As an illustrative example, it could be linked for example with neuromodulation (ie dopamine, see Benchenane et al, 2010) which itself causally modifies the choice process. In this case, coherence would be an excellent predictor of accuracy (as the authors show) without implying a causally important information exchange between the two regions, since inhibiting these regions is without effect.

Altogether, this novel paradigm provides finer control to analyse the role of theta coherence in behaviour. This allows pinpointing of interesting cases in which coherence increases during correct task performance although it may have at least an indirect causal role. This opens up the possibility of interrogating when inter-area synchrony is associated with information transfer and when this information is then used to drive behavioural decisions.

Author Response

The following is the authors’ response to the original reviews.

Reviewer #1 (Recommendations For The Authors):

The brain-machine interface used in this study differs from typical BMIs in that it's not intended to give subjects voluntary control over their environment. However, it is possible that rats may become aware of their ability to manipulate trial start times using their neural activity. Is there any evidence that the time required to initiate trials on high-coherence or low-coherence trials decreases with experience?

This is a great question. First, we designed the experiment to avoid this possibility. Rats were experienced on the sequence of the automatic maze both pre and post implantation (totaling to weeks of pre-training and habituation). As such, the majority of the trials ever experienced by the rat were not controlled by their neural activity. During BMI experimentation, only 10% of trials were triggered during high coherence states and 10% for low coherence states, leaving ~80% of trials not controlled by their neural activity. We also implemented a pseudo-randomized trial sequence. When considered together, we specifically designed this experiment to avoid the possibility that rats would actively use their neural activity to control the maze.

Second, we had a similar question when collecting data for this manuscript and so we conducted a pilot experiment. We took 3 rats from experiment #1 (after its completion) and we required them to perform “forced-runs” over the course of 3-4 days, a task where rats navigate to a reward zone and are rewarded with a chocolate pellet. The trajectory on “forced-runs” is predetermined and rats were always rewarded for navigating along the predetermined route. Every trial was initiated by strong mPFC-hippocampal theta coherence. We were curious as to whether time-to-trial-onset would decrease if we repeatedly paired trial onset to strong mPFC-hippocampal theta coherence. 1 out of 3 rats (rat 21-35) showed a significant correlation between time-to-trial onset and trial number, indicating that our threshold for strong mPFC-hippocampal theta coherence was being met more quickly with experience (Figure R1A). When looking over sessions and rats, there was considerable variability in the magnitude of this correlation and sometimes even the direction (Figure R1B). As such, the degree to which rat 21-35 was aware of controlling the environment by reaching strong mPFC-hippocampal theta coherence is unclear, but this question requires future experimentation.

Strong mPFC-hippocampal theta coherence was used to control trial onset for the entirety of forced-navigation sessions.

Time-to-trial onset is a measurement of how long it took for strong coherence to be met. A) Time-to-trial onset was averaged across sessions for each rat, then plotted as a function of trial number (within-session experience on the forced-runs task). Rat 21-35 showed a significant negative correlation between time-to-trial onset and trial number, indicating that time-to-coherence reduced with experience. The rest of the rats did not display this effect. B) Correlation between trial-onset and trial number (y-axis; see A) across sessions (x-axis). A majority of sessions showed a negative correlation between time-to-trial onset and trial number, like what was seen in (A), but the magnitude and sometimes direction of this effect varied considerably even within an animal.

Is there any evidence that rats display better performance on trials with random delays in which HPC-PFC coherence was naturally elevated?

This question is now addressed in Extended Figure 5 and discussed in the section titled “strong prefrontal-hippocampal theta coherence leads to correct choices on a spatial working memory task”.

The introduction frames this study as a test of the "communication through coherence" hypothesis. In its strongest form, this hypothesis states that oscillatory synchronization is a pre-requisite for inter-areal communication, i.e. if two areas are not synchronized, they cannot transfer information. Recent experimental evidence shows this relationship is more likely inverted-coherence is a consequence of inter-areal interactions, rather than a cause. See Schneider et al. (DOI: 10.1016/j.neuron.2021.09.037) and Vinck et al. (10.1016/j.neuron.2023.03.015) for a more in-depth explanation of this distinction. The authors should expand their treatment of this hypothesis in light of these findings.

Our introduction and discussions have sections dedicated to these studies now.

Figure 6 - It would be much more intuitive to use the labels "Rat 1", "Rat 2", and "Rat 3"; the "21-4X" identifiers are confusing.

This was corrected in the paper.

Figure 6C - The sub-plots within this figure are rather small and difficult to interpret. The figure would be easier to parse if the data were presented as a heatmap of the ratio of theta power during blue vs. red stim, with each pixel corresponding to one channel.

This suggestion was implemented in the paper. See Fig 6C. Extended Fig. 8 now shows the power spectra as a function of recording shank and channel.

Ext. Figure 2B - What happens during an acquisition failure? Instead of "Amount of LFP data," consider using "Buffer size".

Corrected.

Ext. Figure 2D-E - Instead of "Amount of data," consider using "Window size"

Referred to as buffer size.

Ext. Figure 2E - y-axis should extend down to 4 Hz. Are all of the last four values exactly at 8 Hz?

Yes. Values plateau at 8Hz. These data represent an average over ~50 samples.

Ext. Figure 2F - consider moving this before D/E, since those panels are summaries of panel F

Corrected.

Ext. Figure 4A - ANOVA tells you that accuracy is impacted by delay duration, but not what that impact is. A post-hoc test is required to show that long delays lead to lower accuracy than short ones. Alternatively, one could compute the correlation between delay duration and proportion correctly for each mouse, and look for significant negative values.

We included supplemental analyses in Extended Fig. 4

Reviewer #2 (Recommendations For The Authors):

The authors should replace terms that suggest a causal relationship between PFC-HPC synchrony and behavior, such as 'leads to', 'biases', and 'enhances' with more neutral terms.

Causal implications were toned down and wherever “leads” or “led” remains, we specifically mean in the context of coherence being detected prior to a choice being made.

The rationale for the analysis described in the paragraph starting on line 324, and how it fits with the preceding results, was not clear to me. The authors also write at the start of this paragraph "Given that mPFC-hippocampal theta coherence fluctuated in a periodical manner (Extended Fig. 5B)", but this figure only shows example data from 2 trials.

The reviewer is correct. While we point towards 3 examples in the manuscript now, we focused this section on the autocorrelation analysis, which did not support our observation as we noticed a rather linear decay in correlation over time. As such, the periodicity observed was almost certainly a consequence of overlapping data in the epochs used to calculate coherence rather than intrinsic periodicity.

Shortly after the start of the results section (line 112), the authors go into a very detailed description of how they validated their BMI without first describing what the BMI actually does. This made this and the subsequent paragraphs difficult to follow. I suggest the authors start with a general description of the BMI (and the general experiment) before going into the details.

Corrected. See first paragraph of “Development of a closed-loop…”.

In Figure 2C, as expected, around the onset of 'high' coherence trials, there is an increase in theta coherence but this appears to be very transient. However, it is unclear what the heatmap represents: is it a single trial, single session, an average across animals, or something else? In Figure 3F, however, the increase appears to be much more sustained.

The sample size was rats for every panel in this figure. This was clarified at the end of Fig. 3.

In Figure 2D, it was not clear to me what units of measurement are used when the averages and error bars are calculated. What is the 'n' here? Animals or sessions? This should be made clear in this figure as well as in other figures.

The sample size is rats. This is now clarified at the end of Fig 2.

Describing the study of Jones and Wilson (2005), the authors write: "While foundational, this study treated the dependent variable (choice accuracy) as independent to test the effect of choice outcome on task performance." (line 83) It was not clear to me what is meant by "dependent" and "independent" here. Explaining this more clearly might clarify how the authors' study goes beyond this and other previous studies.

The reviewer is correct. A discussion on independent/dependent variables in the context of rationale for our experiment was removed.

Reviewer #3 (Recommendations For The Authors):

As explained in the public review, my comments mainly concern the interpretation of the experimental paradigm and its link with previous findings. I think modifying these in order to target the specific advance allowed by the paradigm would really improve the match between the experimental and analytical data that is very solid and the author's conclusions.

Concerning the paradigm, I recommend that the authors focus more on their novel ability to clearly dissociate the functional role of theta coherence prior to the choice as opposed to induced by the choice. Currently, they explain by contrasting previous studies based on dependent variables whereas their approach uses an independent variable. I was a bit confused by this, particularly because the task variable is not really independent given that it's based on a brain-driven loop. Since theta coherence remains correlated with many other neurophysiological variables, the results cannot go beyond showing that leading up to the decision it correlates with good choice accuracy, without providing evidence that it is theta coherence itself that enhances this accuracy as they suggest in lines 93-94.

The reviewer is correct. A discussion on independent/dependent variables in the context of rationale for our experiment was removed.

Regarding previous results with muscimol inactivation, I recommend that the authors expand their discussion on this point. I think that their correlative data is not sufficient to conclude as they do that despite "these structures being deemed unnecessary" (based on causal muscimol experiments), they "can still contribute rather significantly" since their findings do not show a contribution, merely a correlation. This extra discussion could include possible explanations of the apparent, and thought-provoking discrepancies that they uncover such as: theta coherence may be a correlate of good accuracy without an underlying causal relation, theta coherence may always correlate with good accuracy but only be causally important in some tasks related to spatial working memory or, since muscimol experiments leave the brain time to adapt to the inactivation, redundancy between brain areas may mask their implication in the physiological context in certain tasks (see Goshen et al 2011).

The second paragraph of the discussion is now dedicated to this.

Possible further analysis :

In Extended 4A the authors show that performance drops with delay duration. It would be very interesting to see this graph with the high coherence / low coherence / yoked trials to see if the theta coherence is most important for longer trials for example.

This is a great suggestion. Due to 10% of trials being triggered by high coherence states, our sample size precludes a robust analysis as suggested. Given that we found an enhancement effect on a task with minimal spatial working memory requirements (Fig. 4), it seems that coherence may be a general benefit or consequence of choice processes. Nonetheless, this remains an important question to address in a future study.

Figure 6: The authors explain in the text that although the effect of stimulation of VMT is variable, overall VMT activation increased PFC-HPC coherence. I think in the figure the results are only shown for one rat and session per panel. It would be interesting to add a figure including their whole data set to show the overall effect as well as the variability.

The reviewer is correct and this comment promoted significant addition of detail to the manuscript. We have added an extended figure (Ext. Fig. 9) showing our VMT stimulation recording sessions. We originally did not include these because we were performing a parameter search to understanding if VMT stimulation could increase mPFC-hippocampal theta coherence. The results section was expanded accordingly.

Changes to writing / figures :

The paper by Eliav et al, 2018 is cited to illustrate the universality of coupling between hippocampal rhythms and spikes whereas the main finding of this paper is that spikes lock to non-rhythmic LFP in the bat hippocampus. It seems inappropriate to cite this paper in the sentence on line 65.

We agree with the reviewer and this citation was removed.

Line 180 when explaining the protocol, it would help comprehension if the authors clearly stated that "trial initiation" means opening the door to allow the rat to make its choice. I was initially unfamiliar with the paradigm and didn't figure this out immediately.

We added a description to the second paragraph of our first results section.

Lines 324 and following: the analysis shows that there is a slow decay over around 2s of the theta coherence but not that it is periodical (as in regularly occurring in time), this would require the auto-correlation to show another bump at the timescale corresponding to the period of the signal. I recommend the authors use a different terminology.

This comment is now addressed above in our response to Reviewer #2.

Lines 344: I am not sure why the stable theta coherence levels during the fixed delay phase show that the link with task performance is "through mechanisms specific to choice". Could the authors elaborate on this?

We elaborated on this point further at the end of “Trials initiated by strong prefrontal-hippocampal theta coherence are characterized by prominent prefrontal theta rhythms and heightened pre-choice prefrontal-hippocampal synchrony”

Line 85: "independent to test the effect of choice outcome on task performance." I think there is a typo here and "choice outcome" should be "theta coherence".

The sentence was removed in the updated draft.

Version 2

Abstract

Abstract

Functional interactions between the prefrontal cortex and hippocampus, as revealed by strong oscillatory synchronization in the theta (6-11 Hz) frequency range, correlate with memory-guided decision-making. However, the degree to which this form of long-range synchronization influences memory-guided choice remains unclear. We developed a brain machine interface that initiated task trials based on the magnitude of prefrontal hippocampal theta synchronization, then measured choice outcomes. Trials initiated based on strong prefrontal-hippocampal theta synchrony were more likely to be correct compared to control trials on both working memory-dependent and -independent tasks. Prefrontal-thalamic neural interactions increased with prefrontal-hippocampal synchrony and optogenetic activation of the ventral midline thalamus primarily entrained prefrontal theta rhythms, but dynamically modulated synchrony. Together, our results show that prefrontal-hippocampal theta synchronization leads to a higher probability of a correct choice and strengthens prefrontal-thalamic dialogue. Our findings reveal new insights into the neural circuit dynamics underlying memory-guided choices and highlight a promising technique to potentiate cognitive processes or behavior via brain machine interfacing.

Introduction

Working memory, the ability to temporarily maintain and mentally manipulate information, is fundamental to cognition (Baddeley, 1986). This ability is known to require communication across distributed brain regions and is conserved over mammalia (Goldman-Rakic, 1991; Sarnthein et al., 1998; Lee and Kesner, 2003; Winter and Stich, 2005; Wang and Cai, 2006; Eichenbaum, 2008; Fell and Axmacher, 2011; Christophel et al., 2017; Eichenbaum, 2017; Churchwell and Kesner, 2011; Spellman et al., 2015; Hallock et al., 2016; Ito et al., 2015; Bolkan et al., 2017; Ito et al., 2018; Maisson et al., 2018; Lugtmeijer et al., 2021). Long-range interactions are thought to be supported by the proper timing of action potentials (spikes), and brain rhythms are thought to act as a clocking mechanism to synchronize the timing of spike discharges (Fries, 2005; Buzsaki, 2006; Fell and Axmacher, 2011; Colgin, 2011; Fries, 2015). Fluctuations in the local field potential (LFP) are coupled to the organization of hippocampal spiking activity in rats (O’Keefe and Recce, 1993), primates (Jutras et al., 2009), and humans (Qasim et al., 2021), although the exact frequency can vary over mammalia. The hypothesis that brain rhythms coordinate brain communication by synchronizing neuronal activity, known as “communication through coherence”, is just beginning to be experimentally tested (Fries, 2005; Buzsaki, 2006; Fell and Axmacher, 2011; Fries, 2015; Reinhart and Nguyen, 2019).

In rats, decades of research have shown that computations within, and communication between, the medial prefrontal cortex (mPFC) and hippocampus are required for spatial working memory (Dudchenko et al., 2000; Lee and Kesner, 2003; Wang and Cai, 2006; Horst and Laubach, 2009; Churchwell and Kesner, 2011; Hallock et al., 2013a). Recording studies specifically implicate theta synchrony within the mPFC-hippocampal network as a mechanism for mPFC-hippocampal communication. One metric of oscillatory synchrony, coherence, has been repeatedly correlated with memory-guided choices (Jones and Wilson, 2005; Benchenane et al., 2010; Sigurdsson et al., 2010; O’Neill et al., 2013; Hallock et al., 2016), but also with attention and task engagement (Guise and Shapiro, 2017; Bygrave et al., 2019). In a cornerstone experiment, Jones and Wilson (2005) showed that 4-12Hz mPFC-hippocampal coherence was stronger before rats made a correct choice when compared to a choice error or a forced navigation trial on a spatial memory task. Importantly, these results are derived from measurements of magnitude squared coherence, a measurement of signal correlation, with no requirement for exact numerical phase consistency. For example, two structures can exhibit strong magnitude squared coherence, despite two signals being approximately 180-degree offset in phase. This is an important distinction because there currently exist two versions of the communication through coherence hypothesis; first, that inter-areal communication varies with signal phase, irrespective of coherence, and second that inter-areal communication varies with coherence (Vinck et al., 2023). Likewise, the finding that mPFC-hippocampal theta coherence was stronger on correct choice outcomes is potentially conflated with the fact that rodent movement behaviors also change with task performance (Redish, 2016). Due to constraints on experimental design, it remains unclear as to whether strong theta coherence increased the likelihood of a correct choice, or whether a correct choice led to stronger theta coherence. Addressing this question is of critical importance for the potential use of oscillatory dynamics in therapeutic settings (Reinhart and Nguyen, 2019).

We hypothesized that if magnitude squared coherence represents a valid mechanism for prefrontal-hippocampal communication, that we could use times of strong mPFC-hippocampal theta coherence to gate access to the choice, and that these trials would be associated with better performance on memory-guided tasks. To circumvent a purely correlational experimental design, we developed programmatic algorithms to define and detect strong and weak oscillatory synchronization, then tied theta (6-11Hz) coherence magnitude with task manipulation. This brain machine interface monitored details about task trials, like delay duration and choice outcome, while dynamically adjusting future trials to serve as within-subject controls. Trials initiated during times of strong mPFC-hippocampal theta coherence were associated with correct choice outcomes on both spatial working memory dependent and independent tasks. In follow-up experiments, we found that mPFC theta rhythms and mPFC-thalamic interactions increased with mPFC-hippocampal theta synchrony. Consistent with these results, optogenetic activation of the ventral midline thalamus, a structure known to coordinate mPFC-hippocampal interactions (Vertes, 2002; McKenna and Vertes, 2004; Gabbot et al., 2005; Vertes et al., 2006; Hoover and Vertes, 2007; Hoover and Vertes, 2012; Hallock et al., 2016; Dolleman-van der Weel et al., 2019; Griffin, 2021), dynamically modulated mPFC and hippocampal theta oscillation power and coherence.

Results

Development of a closed-loop brain machine interface for coherence-dependent task manipulation

Our first objective was to design and implement a brain machine interface that would time the initiation of task trials to periods of strong or weak prefrontal-hippocampal theta synchronization (Fig. 1A, Extended Fig. 1, Extended Fig. 3). To do this, we first trained rats to perform a delayed spatial alternation task in a T-maze until reaching 70% choice accuracy on two consecutive days. On this spatial working memory task, rats are rewarded for alternating between left and right reward zones and sequestered at the base of the maze before each choice (Fig. 1A). The ability of this task to tax working memory was validated by measuring the impact of delay duration on choice outcome. Consistent with the use of delayed-response tasks across species (Dudchenko, 2004; Goldman-Rakic, 1991; Eichenbaum, 2008), longer delay durations were associated with lower choice accuracy (Extended Fig. 4A).

Rats were implanted with stainless steel wires targeting the prelimbic and infralimbic subregions of the medial prefrontal cortex (mPFC) and CA1 of dorsal hippocampus (Figs 1A and 2A) to record local field potentials (LFPs). During training sessions, thousands of theta coherence values were calculated during the delay phases, and distributions of mean theta coherence estimates were created (Extended Fig. 2J). Using these distributions, we defined weak theta coherence as 1std below the mean, and strong theta coherence as 1std above the mean of all theta coherence values. Therefore, each rat had a unique numerical value defining states of strong and weak theta coherence, which we could then use as thresholds to initiate trials on the automated maze. Trial initiation is defined by the lowering of the choice-point door to allow access to the maze (Fig. 1A; Extended Fig. 3).

To support brain machine interfacing (see methods section “Brain machine interface”), we designed two independent loops, one processing the neural data in real time and the other controlling the automatic maze (Extended Fig. 1; Fig. 1A). This closed loop system allowed us to monitor prefrontal-hippocampal theta coherence in real time and on a subset of trials, initiate the start of the trial when coherence was strong or weak. While coherence was being monitored, rats were confined to an area at the base of the maze. Trials were initiated by opening a door, providing access to the maze (Fig. 1A).

A brain machine interface that harnesses endogenous mPFC-hippocampal theta coherence on a working memory task.

A) Schematic of brain machine interfacing as rats performed a delayed alternation task on an automated T-maze. The delayed alternation task requires rats to alternate between left and right reward zones. Blue arrows and stars denote correct (rewarded) trajectories while red arrows and stars represent incorrect (unrewarded) trajectories. The rat was confined to the delay zone with three barriers. On a subset of trials, we computed mPFC-hippocampal theta coherence in real time during the delay and trials were initiated contingent upon theta coherence magnitude. B) Frequency by coherence distribution calculated on data collected in real time. For brain machine interfacing experiments, theta coherence was defined as the averaged coherence values between 6-11Hz. Data are represented as the mean +/-s.e.m. C) Thresholds for high and low magnitude coherence were estimated based on distributions of theta coherence values that were unique to individual rats (see Extended Fig. 2 and methods). N = 8 rats (4 female, 4 male).

Strong prefrontal-hippocampal theta coherence leads to correct choices on a spatial working memory task

Based on multiple reports, mPFC-hippocampal theta coherence is positively correlated with memory-guided decision making (Jones and Wilson, 2005; Benchenane et al., 2010; Hallock et al., 2016), but whether theta coherence can be harnessed to bias choice accuracy remains unexplored. To test this idea, we implemented the algorithms described above with an automatic maze to control trial onset via lowering the door for access to the choice (Fig. 1A; Extended Figs 1 and 3). During experimentation, our brain machine interface was activated as rats occupied the delay zone and rats were presented with various trial types within a given session as follows. A small proportion of trials were initiated when mPFC-hippocampal theta coherence was above the strong theta coherence threshold (∼10% of trials) or below the weak theta coherence threshold (∼10% of trials) (Fig. 2A and 2B). Since increasing delay durations led to worse task performance (Extended Fig. 4A), rats also experienced trials that were yoked to high and low coherence trials via identical delay durations. For example, if trial N was a high coherence trial, our algorithm logged the duration spent in the delay zone to be presented back to the rat within a 10-trial block. Thus, initiation of yoked trials was independent of the strength of theta coherence (Fig. 2C) and by comparing choice accuracy on strong/weak coherence trials to that on yoked trials, we were able to rule out the possible confounding variable of working memory load on choice accuracy.

We predicted that, relative to yoked control trials, trials presented during states of strong mPFC-hippocampal theta coherence would be more likely to be correct and trials presented during states of weak mPFC-hippocampal theta coherence would be more likely to be incorrect. Consistent with our first prediction, presenting trials during elevated states of mPFC-hippocampal theta coherence improved choice accuracy (Fig. 2D). However, choice accuracy on trials presented during states of low mPFC-hippocampal theta coherence did not differ from choice accuracy on yoked control trials, indicating that naturally occurring weak theta synchronization does not impair choice outcomes.

Most task trials (∼80%) were initiated after a random delay, irrespective of the magnitude of mPFC-hippocampal theta coherence. We next analyzed whether random delay trials that were coincident with strong mPFC-hippocampal theta coherence also led to correct choice outcomes. First, compared to brain machine interfacing trials, random delay trials coincident with strong mPFC-hippocampal theta coherence were found to be significantly longer in duration (Extended Fig. 5A; BMI trials: mean = 11.55s, std = 1.51s; Random Trials with Strong Theta Coherence: mean = 15.5s, std = 2.2s), an important finding because task performance is impacted by time spent in the delay (Extended Fig. 4A). Unlike brain machine interfacing trials, which had yoked conditions built into 10-trial blocks to account for changing behavior over time, random delay trials that were triggered during strong mPFC-hippocampal theta coherence states were not programmed to have a control. As such, we approximated a yoked condition by identifying random delay trials with identical delay durations as random delay trials with high theta coherence (Extended Fig. 5C). These trials were distributed throughout the session and were unequal in contribution (i.e. there may exist multiple 7s trials to match a 7s random trial with coincident strong theta coherence). Although there was no significant difference between random delay trials coincident with strong theta coherence compared to trials with identical delay durations (p = 0.059; Extended Fig. 5B), 6/8 animals showed better task performance when mPFC-hippocampal theta coherence was strong (Extended Fig. 5D). Given that this comparison is fundamentally different from the brain machine interfacing experiment due to imbalanced design between estimated yoked trials and random trials with high coherence, and did not account for trials with potential salient/distracting events in the environment, we consider these results consistent with our brain machine interfacing findings.

We then examined various measurements of overt behavior to test if behaviors differed between coherence-triggered trials and yoked trials. First, we examined the amount of time spent until rats made a choice, defined as the amount of time from the point at which a trial is initiated until rats passed the infrared beam that triggers the reward dispenser (Extended Fig. 1). While we found no difference in time-to-choice between high coherence trials and yoked trials, there was a trending difference between low and yoked trials (Extended Fig. 4B). Using an analysis to test head-movement complexity (IdPhi; Papale et al., 2012; Redish, 2016), we found no differences between high coherence trials and yoked trials but did observe less head-movement complexity on low coherence trials relative to yoked trials (Extended Fig. 4C). Next, we analyzed total distance traveled in the epoch used to trigger trials during high and low coherence states (last 1.25s before trial initiation). Since the amount of time was always consistent (1.25s), this approach is a proxy for speed, an indirect correlate of theta frequency (Kropff et al., 2021). We found no differences in movement behavior between coherence trials and yoked trials (Extended Fig. 4D). Finally, we found that rats spent similar amounts of time in the delay zone during high and low coherence trials (Extended Fig. 4E). These analyses show that high coherence trials could be used to promote correct choices in the absence of overt differences in behavior between trial types, indicating that mPFC-hippocampal theta coherence preceding the choice potentially influences choice outcome.

High mPFC-hippocampal theta coherence can be used to enhance performance of a working memory dependent task

A) Left panel: Histology from a representative rat showing electrode tracks in the dorsal hippocampus (top) and mPFC (bottom). Right panel: Distribution of trial-types within a session. Within 10-trial blocks, 20% of trials were initiated based on high or low mPFC-hippocampal theta coherence, 20% of trials were yoked to the high/low coherence trials, and 60% were triggered following a random delay (5-30s). Yoked trials were identical in delay duration as high/low coherence trials, but triggered independent of coherence magnitude to control for the negative correlation between delay length and task performance (Extended Fig. 4A). B) Example LFP traces recorded during high and low coherence trials from three representative rats. The mPFC and hippocampal signals were used to compute theta coherence in real-time. C) Rat-averaged coherograms representing time around trial initiation (x-axis), coherence frequency (y-axis) and coherence magnitude, with warmer colors indicating higher coherence values. White arrows denote strong (top panel) and weak (bottom panel) theta coherence, as expected on trials triggered during high and low coherence states. Notice that on yoked trials, coherence was rather consistent before and after trial initiation, as expected for trials triggered independent of coherence magnitude. D) Relative to yoked trials, presenting choices to rats when mPFC-hippocampal theta coherence was high led to improved task performance (t(7) = 2.85, pp.c. = 0.0248). Trials contingent upon low magnitude theta coherence did not impact task performance compared to delay matched controls (t(7) = -0.26, pp.c. = 0.80; paired t-test). Follow-up statistical testing revealed that choice accuracy on high coherence trials was significantly greater than choice accuracy on random delays, consistent with our planned comparisons between high and yoked trials (t(7) = 6.12; p(x4) = 0.002). See Extended Table 1 for statistics. * p<0.05, **p<0.01. Stars (**) above bar graph denotes significance as measured from comparisons relative to random delay choice outcomes (black) and relative to 70% criterion (gray). Subscript “P.C.” indicates planned comparisons. Subscript “(x4)” indicates unplanned comparisons with Bonferroni corrected p-values for the number of unplanned tests performed. N = 8 rats (4 male, 4 female).

Trials initiated by strong prefrontal-hippocampal theta coherence are characterized by prominent prefrontal theta rhythms and heightened pre-choice prefrontal-hippocampal synchrony

Next, we performed offline data analysis to understand the neural dynamics occurring during the high coherence states that improved spatial working memory task performance. First, we noticed that theta rhythms were better characterized by changes within the 6-9Hz range (Fig. 3A) and as such, offline analyses focused on this narrow band. Relative to low coherence states, mPFC theta rhythms were stronger during high coherence states (Fig. 3A-3B; see Fig. 2B for example LFP traces). Hippocampal theta rhythms only exhibited a modest elevation in theta power relative to low coherence states. With respect to theta frequency, mPFC theta rhythms were shifted towards higher frequencies during high coherence states (mean theta frequency = 5.8Hz) relative to low coherence states (mean theta frequency = 5Hz) (Fig. 3C). While there was no significant difference in hippocampal theta frequency, 6/8 rats showed higher theta frequency during high mPFC-hippocampal theta coherence states (mean theta frequency during high coherence states = 7Hz; mean theta frequency during low coherence states = 6.5Hz). We then analyzed whether these signals exhibited evidence of directionality, the ability for one signal to predict another signal as measured by Granger causality analysis (Cohen, 2014). Relative to low coherence states, high coherence states were characterized by stronger hippocampal-to-mPFC theta directionality (Fig. 3D). Thus, the high theta coherence states used to trigger spatial working memory trials were characterized by strong mPFC theta rhythms and hippocampal-to-mPFC theta directionality.

Even though the delay zone was physically close to the choice point (∼30cm), we wondered whether strong mPFC-hippocampal theta coherence trials impacted synchronization during the goal-arm choice. Therefore, we defined choice-point entry as the infrared beam break immediately preceding the choice (Extended Fig. 1). On average, rats took 1.6s and 2.1s to reach this infrared beam from trial initiation on low and high coherence trials, respectively. No significant difference in time-to-choice was observed between high and low coherence trials (Fig. 3E). Thus, we extracted LFPs from -2s to +0.5s surrounding choice-entry (Fig. 3E), and calculated coherence over time and frequency (Fig. 3F). A normalized difference score was calculated from the resultant coherograms (high-low/high+low), revealing a clear difference in theta coherence magnitude between high and low coherence trials as rats approached the choice zone (Fig. 3G). As expected, high coherence trials showed significantly stronger synchronization at -2s, an approximate for trial initiation (Fig. 3H). Interestingly, after the 2s time-point, theta coherence between high and low coherence trials became more similar, but once again differed at ∼0.4-0.5s pre-choice and post-choice entry (Fig. 3H). This latter result shows that strong mPFC-hippocampal theta coherence during the delay was maintained throughout choice.

High mPFC-hippocampal theta coherence trials are gated by prefrontal theta rhythms and lead to heightened pre-choice synchrony

A) Prefrontal and hippocampal power spectra during the high and low coherence epochs used for brain machine interfacing (Fig. 1 and 2). B) Prefrontal theta power (6-9Hz) was significantly greater during high coherence epochs relative to low coherence epochs (t(7) = 5.3, ci = 0.14 to 0.37, padj(x2) = 0.002). Hippocampal theta power was stronger on high coherence compared to low coherence trials (t(7) = 2.47, ci = 0.006 to 0.28, padj(x2) = 0.08, pnot-adj = 0.0427). C) The frequency of prefrontal theta oscillations was significantly higher during high coherence states relative to low coherence states (PFC: t(7) = 3.08, padj(x2) = 0.036, ci = 0.16 to 1.3; hippocampus: t(7) = 1.8, ci = -.17 to 1.3, p = 0.11). Note that 6/8 rats showed higher theta frequency in the hippocampus on high theta coherence states relative to low theta coherence states. Theta frequency was measured by identifying the frequency corresponding to maximum theta power. D) Hippocampal-to-prefrontal theta directionality was significantly stronger during high theta coherence states relative to low theta coherence states (t(7) = 3.53, ci = [0.12 to 0.64], padj(x3) = 0.029) and was significantly stronger than granger prediction in the prefrontal-to-hippocampal direction (t(7) = 3.33, ci = 0.097 to 0.57, padj(x3) = 0.038). No significant effect was observed in the prefrontal-hippocampal direction (t(7) = 0.909, p = 0.39). E) LFP signals (jittered for visualization) were extracted from 2s before choice point entry (as defined by infrared beam breaks) and 0.5s afterwards. Bar graphs show that the average time to choice-entry for high coherence and low coherence trials was between 1.6-2.1s and did not significantly differ between trial-types (t(7) = 2.0, p = 0.08). F) Averaged coherograms (N = 8 rats) showing coherence as a function of frequency and time surrounding choice point entry. G) Difference of the coherograms shown in F. White arrows point to initial 6-9Hz synchronization at -2s which approximates trial onset (see bar graph in E), and a second time point of heightened theta synchrony before choice entry. H) Normalized difference scores representing theta coherence as a function of time. Theta coherence at choice-entry was significantly stronger on trials triggered by high coherence relative to trials triggered during low coherence (see Extended Table 2 for raw and corrected p-values). Data are represented as the mean +/-s.e.m. across 8 rats. *p<0.05, **p<0.01 paired t-tests with Bonferroni p-value corrections when p<0.05. Difference scores were tested against a null of 0. Magenta lines denote p<0.05 after Benjamini Hochberg corrections.

We observed mPFC-hippocampal theta coherence to fluctuate rhythmically (Extended Figs 2H and 6B), and therefore wondered how predictive past values of mPFC-hippocampal theta coherence were of future values. Using previously collected data (Hallock et al., 2016), we extracted mPFC-hippocampal theta coherence epochs across the duration of a 30s delay on the delayed alternation task from 3 rats (N = 22 sessions; Extended Figs 6A and 7A). We performed an autocorrelation analysis on theta coherence values on a trial-by-trial basis, then compared the results to a temporally shuffled theta coherence distribution. Since we performed a moving window approach (1.25s in 250ms steps), comparisons between real and temporally shuffled coherence estimates were only included after 5 lags (lag #4 relative to 0; Extended Fig. 6C). While theta coherence values were predictive of future theta coherence values, this effect slowly decayed over time, indicating that despite some observations of periodicity, the fluctuations were largely non-periodical (Extended Fig. 6C).

In our brain machine interfacing experiments, trials were initiated when mPFC-hippocampal theta coherence was strong or weak. States of strong mPFC-hippocampal theta coherence increased the probability of a correct choice, while increasing synchronization during task performance. However, when we examined the frequency of strong mPFC-hippocampal theta coherence events when the delay phase was fixed and predictable, strong mPFC-hippocampal theta coherence events did not predict trial initiation (Extended Fig. 6D). When considered with the results above, mPFC-hippocampal theta coherence events predict choice outcome, rather than trial onset.

Prefrontal-hippocampal theta coherence states lead to correct choices on a conditional discrimination task

Our findings from Fig. 2 show that mPFC-hippocampal theta coherence leads to correct spatial working memory-guided choices. We next wondered if this effect was specific to spatial working memory and specifically tested whether strong mPFC-hippocampal theta coherence events were optimal for choices on a task where rats must attend to external stimuli to guide decision making. Rats (N = 3; 1 male, 2 female) were implanted with wires targeting the mPFC and hippocampus (Fig. 4B) and were trained to perform a conditional discrimination task where a floor insert dictated choice outcome (e.g. a wooden floor insert signals a left choice, while a mesh insert signals a right choice; Fig. 4A). This task is similar in difficulty to the delayed alternation task, but requires the dorsal striatum, rather than the hippocampus to perform (Hallock et al., 2013a). Likewise, past research showed that inactivation of the mPFC or the ventral midline thalamus did not disrupt conditional discrimination task performance in well-trained rats (Hallock et al., 2013b, Shaw et al., 2013), indicating that the mPFC-hippocampal network is not required for conditional discrimination task performance. Therefore, we predicted that strong mPFC-hippocampal theta coherence would not improve choice outcomes on this conditional discrimination task.

We collected 35 sessions, of which 16 sessions (7 sessions from 21-48 [male]; 4 sessions from 21-49 [female]; and 5 sessions from 21-55 [female]) met criterion for performance of >70%, alternation of <70%, and a contribution of at least 3 trials. Unexpectedly, we found that initiation of trials during strong mPFC-hippocampal theta coherence enhanced choice accuracy on the conditional discrimination task (Fig. 4C). This finding was surprising given that mPFC-hippocampal theta coherence did not previously correlate with choice outcomes on the conditional discrimination task (Hallock et al., 2016), but consistent with increased mPFC-hippocampal theta coherence on a different cue-guided paradigm (Benchenane et al., 2010). Most importantly, these results show that strong mPFC-hippocampal theta coherence is optimal for decision making behavior regardless of whether working memory and mPFC/hippocampal function is necessary to perform a task.

Trials initiated by strong mPFC-hippocampal theta coherence enhance task performance on a two-choice conditional discrimination task

A) Schematic of the conditional discrimination task. Wooden or mesh floor inserts were used to guide choice behavior. Rats were randomly assigned to insert-reward contingencies. Like the brain machine interfacing experiment on the delayed alternation task, trials were initiated when rats were sequestered in the delay zone. B) Example histology from a representative rat showing electrode placements in the hippocampus and mPFC. C) Trials initiated during high mPFC-hippocampal theta coherence states led to better task performance when compared to yoked control trials (t(15) = 2.23, ci = 0.29 to 12.87, p(p.c.) = 0.04) or when compared to trials triggered following a random delay (t(15) = 3.8, ci = 4.7 to 16.6, p(x2) = 0.002). There was no difference in choice outcome following yoked and random delay trials (t(15) = 1.0, ci = -4.5 to 12.7, p(x2) = 0.33). *p<0.05. **p<0.01. Subscript on p-values show if comparisons were planned (‘p.c.’) or corrected for multiple comparisons (‘x2’). Data are represented as the mean ± s.e.m. N = 16 sessions over 3 rats.

Prefrontal-thalamo-hippocampal network dynamics vary with prefrontal-hippocampal synchronization

So far, we have shown that initiating trials when mPFC-hippocampal theta synchrony is strong leads to correct memory-guided choices. What are the mechanisms supporting strong mPFC-hippocampal theta synchrony leading to improved choice accuracy? Past research showed that mPFC-hippocampal theta synchrony during choice was supported by the ventral midline thalamus (VMT; Hallock et al., 2016). The VMT is anatomically connected with the mPFC and hippocampus (Sesack et al., 1989; Vertes, 2002; McKenna and Vertes, 2004; Vertes, 2006; Hoover and Vertes, 2007; Hoover and Vertes, 2012), providing a source of glutamatergic excitation to both structures (Dolleman-van der Weel et al., 2019). Therefore, we wondered how mPFC-VMT and VMT-hippocampal interactions varied with mPFC-hippocampal theta synchronization.

To probe this question, we examined datasets with simultaneous mPFC, VMT, and dorsal hippocampus recordings from 3 rats performing a spatial working memory task (N = 22/28 sessions; Fig. 5A; Extended Fig. 7B; Stout and Griffin, 2020). We extracted neural data as rats occupied the delay zone, then defined and detected epochs of strong and weak mPFC-hippocampal theta coherence offline (Fig. 5B; Extended Fig. 7A-B). Corroborating the findings from our brain machine interfacing experiment (Figs 2 and 3), high theta coherence states were characterized by strong 6-9Hz theta rhythms in the mPFC (Figs 5C and 5D). Intriguingly, the magnitude change of theta power between high and low coherence states was strongest in the mPFC, followed by the VMT, then the hippocampus (Fig. 5D). Relative to low coherence epochs, the mPFC was differentially and simultaneously synchronized to the VMT and hippocampus during high coherence states (Fig. 5E). Moreover, high coherence states were characterized by a stronger change in neural synchronization between the mPFC and VMT, relative to the VMT and hippocampus (Fig. 5F). This latter result suggested that mPFC-VMT interactions may be particularly sensitive to mPFC-hippocampal synchronization. In support of this conclusion, multivariate granger prediction revealed that mPFC-VMT directionality was elevated during strong relative to weak mPFC-hippocampal theta coherence states (Fig. 5G; middle panel). mPFC-hippocampal directionality was also modulated by mPFC-hippocampal theta coherence magnitude. However, directionality between the VMT and hippocampus was minimally impacted by the magnitude of mPFC-hippocampal theta coherence (Fig. 5G).

Lastly, we examined whether mPFC spike-LFP synchrony was impacted by mPFC-hippocampal theta coherence. Spike-phase entrainment was used to quantify the non-uniformity of spike-phase distributions at theta to measure theta phase locking, and spike-field coherence was used to understand the correlation between spikes and LFP across frequencies (Fig. 5H). Out of 126 mPFC neurons, 46 neurons met criterion for inclusion (see methods). When comparing strong to weak mPFC-hippocampal theta coherence states, there were no significant differences to theta phase entrainment (Fig. 5I) nor to spike-field coherence (Fig. 5J) of mPFC spikes to VMT and hippocampal theta.

We then wondered if strong mPFC-hippocampal theta coherence states modulated the spike timing of a select group of mPFC neurons. During strong mPFC-hippocampal theta coherence states, 8.9% and 7% of mPFC neurons were modulated by hippocampal theta and VMT theta, respectively. This contrasted with weak mPFC-hippocampal theta coherence states, where 4.4% and 2.3% of mPFC neurons were significantly modulated by hippocampal and VMT theta, respectively (Fig. 5K). These findings indicate that the magnitude of mPFC-hippocampal theta synchronization was unrelated to global changes to mPFC spike entrainment to VMT and hippocampal theta rhythms. Instead, relative to low coherence states, states of high mPFC-hippocampal theta coherence were associated with strong mPFC spike phase locking to VMT and hippocampal theta rhythms in a small group of mPFC neurons.

Prefrontal-hippocampal theta synchronization modulates prefrontal-thalamic interactions

A) LFPs were recorded from the mPFC, VMT and hippocampal of 3 rats (N = 22 sessions). Right panel shows triple site recordings taken from a representative rat. Green box shows example tetrode tracks from the mPFC. B) High and low mPFC-hippocampal theta coherence epochs were identified, and LFP from the VMT was extracted. The data shown are collapsed across high or low coherence epochs. C) Frequency by coherence plots from the mPFC (top panel), VMT (middle panel), and hippocampus (bottom panel). Compare these data to Fig. 3. D) Normalized difference scores comparing theta (6-9Hz) power between high and low coherence epochs. There was a main effect of brain region on the coherence difference score (F(2,65) = 20.8; p < 0.001; one-way ANOVA) with each brain area showing higher theta power during high coherence states relative to low coherence states (PFC: p < 0.001; VMT; p < 0.001; HPC: p < 0.001; see Extended Table 3). E) Theta coherence for mPFC-VMT and VMT-HPC was estimated during high and low mPFC-hippocampal theta coherence states. F) mPFC-VMT and VMT-HPC theta coherence was stronger during high when compared to low mPFC-hippocampal theta coherence states. mPFC-VMT theta coherence changed more drastically with mPFC-hippocampal theta coherence magnitude (mPFC-VMT: p < 0.001; VMT-HPC: p < 0.001; mPFC-VMT vs VMT-HPC: p < 0.001; see Extended Table 4). G) Multivariate granger prediction analysis. Left panel shows VMT-HPC theta directionality. Middle panel shows mPFC-VMT theta directionality. Right panel shows mPFC-hippocampal theta directionality. Granger prediction in the mPFC-to-VMT direction was more sensitive to mPFC-hippocampal theta coherence magnitude when compared to granger prediction in the VMT-to-mPFC direction (statistics in Extended Table 5). H) Top panel shows hippocampal LFP (1-sec) and example spikes from an mPFC neuron with significant spike-theta entrainment. Middle panel shows polar plots of the unit in the top panel. Histogram represents the distribution of spike-phase values with the mean result length vector shown as a white bar in the center. Bottom panel shows spike-field coherence for the same neuron. I) Difference score (high-low/high+low) of bootstrapped MRL and Rayleighs Z-statistic for each neuron as a function of hippocampal or VMT theta. No significant differences were found between high and low mPFC-hippocampal theta coherence states. J) Spike-field coherence, represented as a difference score. No effects survived p-value correction. Arrow points to a numerical increase to spike-field coherence at hippocampal 4-6Hz. K) Percentage of significantly modulated mPFC units to VMT theta and hippocampal theta as a function of strong (blue) or weak (red) mPFC-hippocampal theta coherence states. *p<0.05. Data are represented as the mean ± s.e.m.

Optogenetic activation of the VMT dynamically regulates prefrontal-hippocampal theta rhythms

Next, we examined whether artificial theta frequency stimulation of the VMT was sufficient to produce synchronized theta rhythms between the mPFC and hippocampus. To investigate this question, we injected the VMT with AAV5-hSyn-ChR2-eYPF to create and embed channelrhodopsin2 at the membrane of VMT neurons, a light-gated cation channel that promotes excitation of neurons with blue light stimulation (450nm). This injection was combined with simultaneous recordings from the mPFC and hippocampus, as well as a fiber placed in the VMT (Fig. 6A). After 4-6 weeks of recovery to allow for viral expression, we pulsed a blue laser targeting the VMT while recording from the mPFC (N = 3/3 rats) and the hippocampus (N = 2/3 rats; Fig. 6A). As a within-subject control, we also stimulated the VMT with a red laser (638nm). Stimulation with red and blue lasers were randomly interleaved within a recording session and various parameters were explored to identify candidate parameters that would facilitate mPFC-hippocampal coherence.

Optogenetic stimulation of the VMT produced a large negative deflection in the mPFC voltage (Fig. 6D), but reliably increased mPFC oscillation power that closely matched the VMT stimulation frequency across all animals and sessions (Extended Fig. 9). VMT theta rhythm stimulation increased the power of mPFC theta oscillations across all recording channels from a 64ch silicone probe targeting mPFC lamina (Fig. 6B-C; see Extended Fig. 8 as a companion figure to Fig. 6C). Stimulation of the VMT at 7, 15, or 30Hz produced clear changes to the mPFC power spectrum, while 4Hz stimulation was more variable across shanks (Extended Fig. 9H). VMT theta stimulation did not always increase or change hippocampal theta rhythm power, but it often increased or changed the shape of the power spectrum (Extended Figs. 9A-G). Surprisingly, optogenetic activation of the VMT at 7-8Hz was largely disruptive to mPFC-hippocampal theta coherence (Extended Fig. 9A-G), but was nonetheless capable of increasing mPFC-hippocampal theta coherence at unexpected frequencies (Fig. 6E). Specifically, VMT stimulation was better capable of increasing mPFC-hippocampal theta coherence when timed with real-time monitoring of hippocampal oscillations and with sufficient activation. For example, in rat #2, we detected hippocampal oscillation power between 1-50Hz and timed VMT stimulation when 8Hz power was the strongest frequency. This approach increased mPFC-hippocampal theta coherence in the 9Hz band (Fig. 6E). Yet, 7Hz stimulation without timing it to hippocampal oscillatory activity had no effect on mPFC-hippocampal theta coherence (Extended Fig. 9G). Likewise, in rat #3, 7.5Hz stimulation was sufficient to enhance mPFC-hippocampal theta coherence at 8.3-8.4Hz at 4.5mW, but the same was not true at 1mW power (Fig. 6E and Extended Fig. 9A-C).

While we expected VMT stimulation to strengthen mPFC-hippocampal theta coherence, these results indicate that square-wave optogenetic stimulation of the VMT does not pose a viable approach to strengthen mPFC-hippocampal coherence without consideration of on-going oscillatory dynamics. Instead, VMT stimulation most effectively produces closely matched oscillations in the mPFC, a finding with interesting implications for diseases characterized by a disrupted thalamic complex (Elvsåshagen et al., 2021). Future research should perform a systemic characterization of the parameter space and opsins that allow VMT activation with optogenetics to produce mPFC-hippocampal theta coherence. This study is warranted given the growing hypothesis that the VMT regulates mPFC-hippocampal oscillatory dynamics (Dolleman-van der Weel et al., 2019).

Optogenetic activation of the ventral midline thalamus increased prefrontal and hippocampal theta power while dynamically adjusting the mPFC-hippocampal theta coherence distribution

A) Top panel, Schematic demonstrating recordings from the mPFC and hippocampus with optogenetic activation of the VMT. Middle panel, example histological confirmation of fiber implant and viral expression targeting the VMT. Bottom panel. Viral expression at similar viral injection coordinates. Notice that all rats showed overlap in viral expression in the nucleus reuniens (brain section overlay from Paxinos and Watson, 2006). B) Top panel, histological confirmation of 64ch silicon probe recordings in the dorsal medial prefrontal cortex. Bottom panel, Optogenetic activation of the VMT at 7Hz produced prefrontal theta rhythms (N = 83 blue, 88 red laser events; rat #1). C) Ratio of log-transformed mPFC theta (6-9Hz) power between blue and red laser events across silicon probe shanks and channels. Values > 1 indicate that theta during blue laser epochs was stronger than during red laser epochs. “NaN” represents an excluded channel. See Extended Fig. 8 for companion figure. Columns represent recording channels per shank, while rows represent shank number from the corresponding medial-lateral placement in the mPFC (B). D) Data from rat #2 (N = 108 blue, 104 red laser events) and rat #3 (N = 113 blue, 101 red laser events). Top panel shows raw LFP traces, middle panel shows theta filtered traces (6-9Hz), while the bottom panel shows theta coherence as a function of time. Yellow box shows the stimulation event. Arrows point to observed negative deflects in the LFP signals surrounding VMT stimulation onset. E) Power and coherence analyses performed on data during VMT stimulation (0-1.5s from laser onset) as a function of frequency (x-axis), brain region (row), and rat (left/right panels). Both mPFC and hippocampal theta power were increased during VMT stimulation. Coherence between mPFC and hippocampal theta rhythms were reduced or enhanced in a frequency-dependent manner during VMT stimulation. Magenta lines denote p<0.05 following Benjamini-Hochberg p-value corrections for two-sample t-tests between the 6-11Hz range. Data are represented as the mean ± s.e.m.

Discussion

Previous research showed that mPFC-hippocampal theta coherence was stronger when memory was used to guide choices (Jones and Wilson, 2005; Benchenane et al., 2010; Sigurdsson et al., 2010; O’Neill et al., 2013; Hallock et al., 2016), but this conclusion required correlating choice outcome with mPFC-hippocampal theta synchrony. Unlike past work, we manipulated the timing of trial-onset relative to the strength of mPFC-hippocampal theta synchrony and as such, the detection of coherence state always preceded choice outcome. Our brain machine interfacing experiments allowed us to implement various within subject controls and we showed that trials initiated during states of strong mPFC-hippocampal theta coherence led to better task performance on two separate paradigms.

While we expected this form of long-range theta synchronization to be particularly useful when spatial working memory was used to guide decision-making, we also observed that mPFC-hippocampal theta coherence enhanced the performance of a task that did not require the mPFC, VMT, nor hippocampus for successful performance (Hallock et al., 2013a; Hallock et al., 2013b, Shaw et al., 2013). These findings raise an interesting discrepancy - mPFC-hippocampal theta coherence led to improved task performance on the conditional discrimination task, yet pharmacological inactivation of these structures did not impair task performance. Given that the conditional discrimination task is dependent on the dorsal striatum, it is possible that pharmacological techniques, which work on the scale of minutes, provided time for the brain to adapt to a disrupted mPFC-hippocampal network. In support of this view, Goshen and colleagues (2011) showed that optogenetic suppression of the CA1 on a time-scale similar to pharmacological agents, like muscimol, did not impair the retrieval of a contextual fear memory. However, when optogenetic inactivation was temporally specific to the testing phase of the contextual fear memory paradigm, memory retrieval was disrupted. These findings show that the timescale of inactivation impacts the results and conclusions drawn from research, raising the possibility that the mPFC-hippocampal network can indeed be beneficial to the performance of working memory-independent tasks. Future research should be dedicated to testing the causal link of mPFC-hippocampal theta synchronization to choice outcome by implementing procedures similar to what is described here with optogenetic perturbations.

To then characterize the neural dynamics co-occurring with strong mPFC-hippocampal theta coherence events, we tested whether mPFC-thalamic and hippocampal-thalamic interactions changed with strong and weak mPFC-hippocampal theta synchronization events. For these analyses, we focused on the ventral midline thalamus (VMT), a structure that is bidirectionally connected with the mPFC and hippocampus and supports mPFC-hippocampal neuronal interactions (Vertes, 2002; McKenna and Vertes, 2004; Vertes et al., 2006; Hoover and Vertes, 2007; Hoover and Vertes, 2012; Ito et al., 2015; Hallock et al., 2016). Consistent with mPFC-hippocampal theta coherence reflecting heightened neural coordination across the brain, VMT theta rhythms showed stronger coherence to mPFC and hippocampal theta rhythms when the mPFC and hippocampus were strongly coherent. Likewise, optogenetic activation of the VMT modulated mPFC and hippocampal theta rhythms, while dynamically altering the way in which these structures were coherent at theta. It should be noted that because hippocampal theta rhythms were already prominent, the effect of VMT stimulation could have appeared less dramatic for hippocampal theta relative to mPFC theta. Nonetheless, our physiological and optogenetic work point towards cortico-thalamic dialogue as a central component of mPFC-hippocampal theta synchronization. Importantly, this latter assertion is supported by anatomy, as the mPFC receives no direct projections from the dorsal hippocampus (Jay and Witter, 1991; Hoover and Vertes, 2007), but influences hippocampal neuronal activity via the thalamus (Ito et al., 2015). We suspect that the VMT may coordinate mPFC-hippocampal neural interactions through cortico-thalamo-cortical looping mechanisms, as the VMT projects directly to entorhinal cortex neurons that target the CA1 (Wouterlood, 1991) and modulates CA1 neurons with concurrent cortical activation (Dolleman-Van der Weel, 2017). Consistent with this hypothesis, mediodorsal thalamus is known to sustain mPFC neuronal activity (Bolkan et al., 2017; Schmitt et al., 2017), and the VMT supports mPFC firing and mPFC-hippocampal synchronization (Hallock et al., 2016; Jayachandran, Viena et al., 2023).

If mPFC-hippocampal oscillatory synchronization structured cortico-thalamic and cortico-hippocampal neuronal communication, then we would have expected strong theta synchronization events to correlate with mPFC spike entrainment to hippocampal and VMT theta rhythms. When examining all mPFC neurons, we found no differences to spike-LFP synchronization between strong and weak mPFC-hippocampal theta coherence events. Instead, we found a rather small increase to the percentage of theta modulated units in the mPFC. This observation is consistent with recent experimental, modeling, and theoretical work, implicating coherence as a product of communication, rather than a scaffold (Schneider et al., 2021; Vinck et al., 2023). For example, Schneider and colleagues (2021) showed that LFP signal coherence between a sending and receiving structure can be explained by a sending structures signal power and strength of projectors and can emerge without changes to spike-entrainment in the receiving structure. Given that the ventral hippocampus and ventral midline thalamus are necessary for mPFC-hippocampal theta coherence (O’Neill et al., 2013; Hallock et al., 2016), we suspect that afferents from these structures contribute significantly to mPFC-hippocampal oscillatory synchronization. When taken together, mPFC-hippocampal theta coherence events may represent short temporal periods of neural communication, rather than scaffolding communication. As such, the existing literature combined with our findings strengthen a claim for using patterns of oscillatory synchronization in a therapeutic setting.

Consistent with our work, a recent study found that inducing states of theta synchrony between frontal and temporal regions via transcranial alternating-current stimulation, rescued age-related memory impairments in human participants (Reinhart and Nguyen, 2019). Our findings suggest that tapping into pre-existing neural dynamics holds significant promise for improving memory. We hypothesize that non-invasive stimulation techniques prior to therapy, paired with synchrony-dependent attention or working memory practice via brain machine interfacing, could pose a viable intervention to improve cognitive deficits. In closing, the use of brain machine interfacing holds significant promise for clinical and neuroscientific advance.